The Evolution of Phenotypically Invariant Genetic Networks

### 2017-12-04 kalman_walk_paper.tex
note on inheritance

Each genome an organism inherits is generated by meiosis,
in which both of its diploid parents recombine their two genomes.
We will assume that each coefficient (i.e., entry of $A$, $B$ or $C$)
is determined by a single nonrecombining locus,
so that each coefficient in the system produced by meiosis is an independent random choice
between the two parental coefficients.
With these definitions, an $F_1$ offspring
carries a system copy from each parent,
and an $F_2$ is an offspring of two independently formed $F_1$s.
If the parents are from distinct populations,
these are simply first-- and second--generation hybrids, respectively.

This is a simplification:
since the $i^\text{th}$ row of $A$ summarizes how each gene regulates gene $i$,
and hence is determined by the promoter region of gene $i$,
we would actually expect the elements of a row of $A$ to tend to be inherited together.
Similarly, we expect in practice heritable variation in each coefficient
to be determined by more than one locus --
but this may be a reasonable approximation.

Offspring formed from two phenotypically identical systems do not necessarily exhibit the same phenotype as both of its parents -- in other words $\allS$, the set of all systems phenotypically equivalent to a given one, is not, in general, closed under averaging or recombination.
Next we discuss how this fact can contribute to hybrid incompatibility and genetic load.
If sexual recombination among systems drawn from $\allS$ yields systems with divergent phenotypes, populations containing significant diversity in $\allS$ can carry genetic load, and isolated populations may fail to produce hybrids with viable phenotypes.

@@

\subsection*{Hybrid incompatibility}
Two parents with the optimal phenotype can produce offspring whose phenotype is suboptimal
if the parents have different underlying systems.
%This leads to the question: 
How quickly do hybrid phenotypes break down as genetic distance between parents increases?
To quantify this, we will measure how far a system's phenotype is from optimal
using a weighted difference between impulse response functions.
Suppose that $\rho(t)$ is a nonnegative, smooth, square-integrable weighting function,
suppose that $h_0(t)$ is the \emph{optimal} impulse response function 
and define the ``distance to optimum'' of another impulse response function
to be
\begin{align}
\label{eqn:distance}
	D(h) = \left( \int_0^\infty \rho(t) \|h(t) - h_0(t)\|^2 dt \right)^{1/2} .
\end{align}
Consider reproduction between a parent with system $(A, B, C)$ 
and another displaced by distance $\epsilon$ in the direction $(X,Y,Z)$,
i.e., having  system $(A + \epsilon X, B + \epsilon Y, C + \epsilon Z)$.
We assume both are ``perfectly adapted'' systems, 
i.e., having impulse response function $h_0(t)$,
and their offspring has impulse response function $h_\epsilon(t)$.
A Taylor expansion of $D(h_\epsilon)$ in $\epsilon$ is explicitly worked out in Appendix \ref{apx:H_calc},
and shows that the phenotype of an $F_1$ hybrid between these two is at distance proportional to $\epsilon^2$ from optimal,
while $F_2$ hybrids are at distance proportional to $\epsilon$.
This is because an $F_1$ hybrid has one copy of each parental system,
and therefore lies directly between the parental systems (see Figure \ref{fig:conceptual_fig}) --
the parents both lie in $\allS$, which is the valley defined by $D$,
and so their midpoint only differs from optimal due to curvature of $\allS$.
In contrast, an $F_2$ hybrid may be homozygous for one parental type in some coefficients
and homozygous for the other parental type in others;
this means that each coefficient of an $F_2$ may be equal to either one of the parents,
or intermediate between the two;
this means that possible $F_2$ systems may be as far from the optimal set, $\allS$,
as the distance between the parents.
The precise rate at which the phenotype of a hybrid diverges depends on the geometry --
in Figure \ref{fig:conceptual_fig}, this is depicted as the angle of the black line (the optimal set)
with respect to the coordinates.

>>>>

Each genome an organism inherits is generated by meiosis,
in which both of its diploid parents recombine their two genomes,
and so an $F_1$ offspring carries one system copy from each parent,
and an $F_2$ is an offspring of two independently formed $F_1$s.
If the parents are from distinct populations,
these are simply first-- and second--generation hybrids, respectively.

Exactly how the coefficients 
(i.e., entries of $A$, $B$ or $C$)
of a haploid system inherited by an offspring from her diploid parent
are determined by the parent's two systems
depends on the genetic basis of any variation in the coefficients.
Since the $i^\text{th}$ row of $A$ summarizes how each gene regulates gene $i$,
and hence is determined by the promoter region of gene $i$,
the elements of a row of $A$ to tend to be inherited together,
which will create covariance between entries of the same row.
It is, however, a quite general observation that the variation seen among recombinant systems
is proportional to the difference between the two parental systems.
This is certainly true if each coefficient is determined by a single nonrecombining locus,
so that each coefficient in the system produced by meiosis is an independent random choice
between the two parental coefficients.
On the other hand, if a coefficient is determined by additive contributions of independently segregating loci,
then the law of total variance applied after conditioning on the parental origin of each allele
implies that the variance is equal to one-quarter of the squared difference between parental coefficients.

\plr{continue editing here}

Offspring formed from two phenotypically identical systems do not necessarily exhibit the same phenotype as both of its parents -- in other words $\allS$, the set of all systems phenotypically equivalent to a given one, is not, in general, closed under averaging or recombination.
If sexual recombination among systems drawn from $\allS$ yields systems with divergent phenotypes, populations containing significant diversity in $\allS$ can carry genetic load, and isolated populations may fail to produce hybrids with viable phenotypes.

@@

\subsection*{Hybrid incompatibility}
Two parents with the optimal phenotype can produce offspring whose phenotype is suboptimal
if the parents have different underlying systems.
%This leads to the question: 
How quickly do hybrid phenotypes break down as genetic distance between parents increases?
To quantify this, we will measure how far a system's phenotype is from optimal
using a weighted difference between impulse response functions.
Suppose that $\rho(t)$ is a nonnegative, smooth, square-integrable weighting function,
$h_0(t)$ is the \emph{optimal} impulse response function,
and define the ``distance to optimum'' of another impulse response function
to be
\begin{align}
\label{eqn:distance}
	D(h) = \left( \int_0^\infty \rho(t) \|h(t) - h_0(t)\|^2 dt \right)^{1/2} .
\end{align}
Consider reproduction between a parent with system $(A, B, C)$ 
and another displaced by distance $\epsilon$ in the direction $(X,Y,Z)$,
i.e., having  system $(A + \epsilon X, B + \epsilon Y, C + \epsilon Z)$.
We assume both are ``perfectly adapted'' systems, 
i.e., having impulse response function $h_0(t)$,
and their offspring has impulse response function $h_\epsilon(t)$.
A Taylor expansion of $D(h_\epsilon)$ in $\epsilon$ is explicitly worked out in Appendix \ref{apx:H_calc},
and shows that the phenotype of an $F_1$ hybrid between these two is at distance proportional to $\epsilon^2$ from optimal,
while $F_2$ hybrids are at distance proportional to $\epsilon$.
This is because an $F_1$ hybrid has one copy of each parental system,
and therefore lies directly between the parental systems (see Figure \ref{fig:conceptual_fig}) --
the parents both lie in $\allS$, which is the valley defined by $D$,
and so their midpoint only differs from optimal due to curvature of $\allS$.
In contrast, an $F_2$ hybrid may be homozygous for one parental type in some coefficients
and homozygous for the other parental type in others;
this means that each coefficient of an $F_2$ may be equal to either one of the parents,
or intermediate between the two;
this means that possible $F_2$ systems may be as far from the optimal set, $\allS$,
as the distance between the parents.
The precise rate at which the phenotype of a hybrid diverges depends on the geometry --
in Figure \ref{fig:conceptual_fig}, this is depicted as the angle of the black line (the optimal set)
with respect to the coordinates.
